---
title: 'An Analysis of the Breweries and their Beers in the United States'
author: "Caroll Rodriguez and Aaron Tomkins"
date: "2/16/2018"
output: 
    html_document:
      keep_md: true
---
### 
#### The following code chunk pulls in two .csv files provided by the client into R. The files are then merged the Brewery Id Primary key. A sample of the data is shown below.

```{r tidyData, echo=TRUE}
# load Beers.csv, setting data types
classProfile <- c("character", "integer", "numeric", "integer", "integer", "factor", "numeric")
beers <- read.csv("Beers.csv", colClasses = classProfile, na.strings = "")
# further specify Name column in beers
names(beers)[names(beers) == "Name"] <- "Beer_Name"

# load Breweries.csv, setting data types
classProfile <- c("integer","character", "character", "factor")
breweries <- read.csv("Breweries.csv", colClasses = classProfile, na.strings = "")
# further specify Name column in breweries
names(breweries)[names(breweries) == "Name"] <- "Brewery_Name"

master <- merge(beers, breweries, by.x="Brewery_id", by.y="Brew_ID", all=TRUE)
# reorder columns
master <- master[c("Beer_ID", "Beer_Name", "ABV", "IBU", "Style", "Ounces", "Brewery_id", "Brewery_Name", "City", "State" )]
# normalize column names
names(master)[names(master) == "Brewery_id"] <- "Brewery_ID"
# sort rows by beer ID
master <- master[order(master$Beer_ID),]

# head and tail of master
head(master, 6)
tail(master, 6)
```

# Questions

1. How many breweries are present in each state?
<br /><b>The number of breweries in each state is given in the table below.</b>
```{r, echo=TRUE}
table(master$State, useNA = "no")
```

2. Merge beer data with the breweries data. Print the first 6 observations and the last six observations to check the merged file.
<br /><b>The first and last six observations are printed below.</b>
```{r, echo=TRUE}
# The two tables were already merged in the data cleaning process (see first section above).
head(master, 6)
tail(master, 6)
```

3. Report the number of NA's in each column.
<br /><b>Per below, the ABV column has 62, the IBU column has 1005, and the Style column has 5 NAs. All other columns have zero.</b>
```{r, echo=TRUE}
sapply(master, function(x) sum(is.na(x)))
```

4. Compute the median alcohol content and international bitterness unit for each state. Plot a bar chart to compare.
<br /><b>The bar chart with the median ABV and IBU per state is given below.</b>
```{r, echo=TRUE}
ibu.med <- aggregate(master$IBU, by = list(master$State), FUN = function(x) median(x, na.rm = TRUE))
names(ibu.med) <- c("State","IBU")
abv.med <- aggregate(master$ABV, by = list(master$State), FUN = function(x) median(x, na.rm = TRUE))
names(abv.med) <- c("State","ABV")

barplot(ibu.med$IBU, names.arg = ibu.med$State, beside = TRUE, main = "Median IBU by State", xlab = "State", ylab = "Median IBU", col = "Blue")
barplot(abv.med$ABV, names.arg = abv.med$ABV, beside = TRUE, main = "Median ABV by State", xlab = "State", ylab = "Median ABV", col = "Blue")
```

5. Which state has the maximum alcoholic (ABV) beer? Which state as the most bitter (IBU) beer?
<br /><b>Colorado has the maximum ABV and Oregon has the most bitter beer.</b>
```{r}
master[which.max(master$ABV),][,10]
master[which.max(master$IBU),][,10]
```

6. Summary statistics for the ABV variable.
```{r}
summary(master$ABV)
```

7. Is there an apparent relationship between the bitterness of the beer and its alcoholic content? Draw a scatter plot.
<br /><b>Per below, there is a positive linear relationship between ABV and IBU.</b>
```{r}
plot(master$ABV ~ master$IBU, main = "ABV to IBU", ylab = "ABV", xlab = "IBU")
```
